### 2021-06-26 Manuscript.Rmd
manuscript edits, repo cleanup

readr::write_csv(edges_tidy_directional_export, "Supplementary_Data/Supplementaryfile6.csv", na = "NA")

@@

# Evolution of gene expression across species and specialized zooids in Siphonophora {-}
Catriona Munro^1,2^,  Stefan Siebert^3^, Mark Howison^4^, Felipe Zapata^5^, Casey W. Dunn^6^
\newline
\newline
^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI 02912, USA \newline
^2^ Current address: Collège de France, PSL Research University, CNRS, Inserm, Center for Interdisciplinary Research in Biology, 75005 Paris, France \newline
^3^ XXXXXX Insert Stefan address XXXXXXXXXXX \newline
^4^ Research Improving People's Lives (RIPL), Providence, RI, USA \newline
^5^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA 90095, USA \newline
^6^ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT 06520, USA

@@

Siphonophores are highly complex colonial organisms, consisting of asexually-produced bodies (called zooids) that share a common gastrovascular cavity but that are functionally specialized to specific tasks, including feeding, swimming, and reproducing. Though this extreme functional specialization has captivated biologists for generations, the genomic underpinnings of this phenomenon remain poorly understood. We use RNA-seq to investigate gene expression patterns in five zooids, and one specialized tissue (the pneumatophore, a gas filled float), across seven siphonophore species. To overcome the analytical challenges of studying the evolution of gene expression across genes and species, we introduce a novel expression metric and two new analytical approaches.

@@

Within species, we identified hundreds of zooid-specific genes, as well as a number of putative transcription factors showing higher abundance in particular zooids and developmental stages. In species with unique zooid types, we found significant differences in expression profiles between zooids with different function (i.e., feeding vs. tentacle in Physalia physalis), but no differences between distinct zooids with similar functions (e.g., feeding zooids in Bargmannia elongata and Agalma elegans). Across all species, we found that gene expression patterns tended to be largely consistent among species rather than zooids, and found evidence of large lineage specific shifts in gene expression.

@@

Our findings show that there are distinct expression patterns defining each functionally specialized zooid or tissue. In species-specific instances of zooid diversification, we could distinguish novel zooids with distinct morphological and functional differences, but were unable to gain insights into differences between zooids with small differences in morphology or colony placement. Using comparative approaches, we also identified shared zooid-specific gene expression patterns across species, and within a given zooid type, identified instances of clade/species-specific expression patterns.

## Introduction {-}

@@

In this study we use RNA-seq data investigate the functional specialization and evolution of zooids across siphonophore species. To carry out these analyses, we also address three key challenges to study the evolution of gene expression in a comparative framework [@dunn2013phylogenetic]. First, we introduce a new metric to normalize gene expression which is valid for comparison across species. A common metric used to quantify gene expression is Transcripts Per Million (TPM) [@li2011rsem; @wagner2012measurement]. TPM is a relative measure of expression which depends on the number of genes present in a reference. Hence, TPM values are a valid metric when comparing libraries within a single species, but are not directly comparable across species when the references are incomplete, or genes have been gained and lost in the course of evolution. To address this issue, we introduce Transcripts Per Million 10K (TPM10K), a metric which normalizes TPM to account for different sequencing depths among species (see Methods for details).

Second, we further account for gene and species effects to compare gene expression across genes and species. Normalized read counts produced by RNA-seq experiments are proportional to gene expression, but they are also impacted by factors that can vary across genes and species such as gene sequence and length. Therefore, to estimate and compare expected gene counts it is necessary to model such factors with unknown species and gene-specific counting-efficiency coefficients [@dunn2013phylogenetic]. A direct comparison of expected gene counts among species, however, can be misleading if differences in counts are simply due to differences in counting efficiency and not due to differences in expression. To address this issue, we use ratios of expected counts to compare expression across genes and species. Using ratios of counts, we are able to eliminate species- and gene- specific counting efficiency within species, as the unknown counting efficiency factor is in both the numerator and the denominator, and is thus removed prior to comparisons among species [@dunn2013phylogenetic].

Third, we use a novel approach to investigate the evolution of gene expression on gene trees with complex evolutionary histories. Most evolutionary studies of gene expression focus exclusively on comparing expression values of strict orthologs (i.e., gene lineages related only by speciation events) which are shared across all species. We call this type of classical analysis Strict Ortholog Filtering of Trees (SOFT). This limits analyses to the subset of genes that have no evidence of duplication. This is an all or nothing approach, where gene trees are retained if they have one sequence for each species, and discarded if there are multiple sequences in any of the species. Because this is usually such a small fraction of genes, traditional SOFT analyses discard a large fraction of the data, including many gene families that may be of interest to the investigator.

Because the history of genes is characterized by more complex scenarios involving gene duplication and speciation, we introduce a new method that takes advantage of this rich history to examine the evolution of gene expression across more genes and species. We call this method Species Tree Edge Extraction and Layover (STEEL) (Fig. \@ref(fig:methods-comp)), as we map expression data to gene phylogenies and identify equivalent branches in the species tree which are descended from speciation events in order to make valid comparisons across species. While SOFT is a gene tree filtering approach, where entire trees are discarded due to even a single duplication event, STEEL is a branch filtering approach. This means that STEEL retains many informative branches from trees that would be removed by SOFT, and preserves many more evolutionary comparisons for analyses.

Briefly, the principles of STEEL are as follows: first we identify speciation and duplication events in a given gene tree, specifically labelling nodes that correspond to speciation events in the species tree (Fig. \@ref(fig:methods-comp), step 1). Next, we map gene expression values to the tips of the gene tree (Fig. \@ref(fig:methods-comp), step 2), and use phylogenetic methods to reconstruct expression values at the internal nodes (Fig. \@ref(fig:methods-comp), step 3). Expression values are mapped and reconstructed for each zooid/tissue separately, with the assumption that the structure is homologous across species. Then, we calculate scaled expression values across branches, by subtracting the expression value at the child node from the parent node, and dividing by branch length (branch lengths are calibrated against branches in the species tree) (Fig. \@ref(fig:methods-comp), step 4). Finally, we identify branches in the gene trees that correspond to equivalent branches in the species tree (hereafter species-equivalent branches) (Fig. \@ref(fig:methods-comp), step 5). Species-equivalent branches are gene tree branches with parent and child nodes that are both speciation events and that correspond to branches in the species tree. Each branch in the species tree is given a unique identifier (i.e., a letter) and the species-equivalent branches in gene trees are given the same identifier (Fig. \@ref(fig:methods-comp), step 5). This method enables the selection of branches within gene trees that are equivalent to branches within the species tree, and are thus comparable with one another across all gene trees. Unlike the SOFT approach, this approach considers equivalent branches that are descended from speciation events, but that have more complex evolutionary histories. For example, due to deeper gene duplication events, gene trees often contain multiple branches that correspond to the same branch in the species tree (Fig. \@ref(fig:methods-comp), step 5). Our method allows us to consider all of these branches.

Using the methods described above, combined with classical differential gene expression analysis, we conducted differential expression analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. Second, we examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Third we compared gene expression across the seven siphonophore species, to identify putative clade/lineage or zooid specific expression patterns.

(ref:methods-comp) New phylogenetic approach to identify evolutionary changes in gene expression, called Species Tree Edge Extraction and Layover (STEEL). Step 1, we label each of the nodes in the species tree, and identify equivalent speciation nodes across every gene tree (an exemplar is shown here). Step 2, we map expression values (TPM10K) to the tips (expression values are mapped and reconstructed for each homologous zooid separately). Step 3, we reconstruct ancestral trait expression values at all internal nodes where expression data are available. Step 4, we calculate scaled change in gene expression (child node expression - parent node expression / branch length). Branch length is calibrated to the species tree branch lengths. Step 5, we identify branches in gene trees that correspond to equivalent branches in the species tree. There may be more than one branch in a gene tree that corresponds to the same branch in the species tree.

@@

# Bargmannia_GO_Gasmat<- Go_lookup("Bargmannia",Bargmannia_Gaswhimat)
# Agalma_GO_Gasmat<- Go_lookup("Agalma",Agalma_gasmat)
# Nanomia_GO_Gasmat<- Go_lookup("Nanomia",Nanomia_Gasmat)
# Frillagalma_GO_Gasmat<- Go_lookup("Frillagalma",Frillagalma_Gasmat)
# Physalia_GO_Gasmat<- Go_lookup("Physalia",Physalia_Gasmat)
# Physalia_GO_Tenpalmat<- Go_lookup("Physalia",Physalia_Tenpalmat)
# 
# Bargmannia_GO_Gasdev<- Go_lookup("Bargmannia",Bargmannia_Gasdev)
# Nanomia_GO_Gasdev<- Go_lookup("Nanomia",Nanomia_Gasdev)
# Frillagalma_GO_Gasdev<- Go_lookup("Frillagalma",Frillagalma_Gasdev)
# Physalia_GO_Gasdev<- Go_lookup("Physalia",Physalia_Gasdev)
# 
# Bargmannia_GO_Gonmal<- Go_lookup("Bargmannia",Bargmannia_Gonmal)
# Frillagalma_GO_Gonmal<- Go_lookup("Frillagalma",Frillagalma_Gonmal)
# Agalma_GO_Gonmal<- Go_lookup("Agalma",Agalma_Gonmal)
# 
# Frillagalma_GO_Gonfem<- Go_lookup("Frillagalma",Frillagalma_Gonfem)
# Agalma_GO_Gonfem<- Go_lookup("Agalma",Agalma_Gonfem)
# Nanomia_GO_Gonfem<- Go_lookup("Nanomia",Nanomia_Gonfem)
# 
# Nanomia_GO_Palmat<- Go_lookup("Nanomia",Nanomia_Palmat)
# Frillagalma_GO_Palmat<- Go_lookup("Frillagalma",Frillagalma_Palmat)
# Agalma_GO_Palmat<- Go_lookup("Agalma", Agalma_Palpon)

# Nanomia_GO_Necdev<- Go_lookup("Nanomia",Nanomia_Necdev)
# Frillagalma_GO_Necdev<- Go_lookup("Frillagalma",Frillagalma_Necdev)
# Agalma_GO_Necdev<- Go_lookup("Agalma", Agalma_necdev)
# Bargmannia_GO_Necdev<-Go_lookup("Bargmannia", Bargmannia_necdev)
# 
# Diphyes_GO_Gon<- Go_lookup("Diphyes",Diphyes_Gon)

@@

We sequenced mRNA from microdissected zooids from seven different siphonophore species (at least two replicate colonies) and mapped these short-read libraries to previously published transcriptomes [@munro2018improved]. We collected RNA-seq data from, where possible, 5 different zooids and one specialized tissue, the pneumatophore, as well as unique zooids specific to *Agalma elegans* (B palpons), *Physalia physalis* (tentacular palpon), and *Bargmannia elongata* (yellow and white gastrozooids), and where possible developing and mature zooids (Table \@ref(tab:summarize-libraries)). Due to species availability, we were not able to sample more than one replicate for some zooids - single replicates were excluded from downstream differential gene expression analyses (see Fig. \@ref(fig:zooids-sampled)).

@@

The second component of variation we considered was biological variation among sampled colonies. Specimens were collected in the wild at different depths and over different time periods, but despite these environmental factors there was remarkably little variation among sampled colonies. Some samples did show greater variation across replicates, such as a developing palpon replicate in *Nanomia bijuga* (Fig. \@ref(fig:pca-nanomia)), and a developing gastrozooid and male gonodendra in *Frillagalma vityazi* (Fig. \@ref(fig:pca-Frillagalma)).

The third component of variation we considered was among zooids/specialized tissues within species. This was the greatest component of variation (Fig. \@ref(fig:pca-diphyes)--\@ref(fig:pca-physalia)). We identified genes that show higher transcript abundance in particular zooids, and in the pneumatophore, for each of the sampled species, based on pairwise comparisons (supplementary file 1). From this, we identified a subset of genes that have higher transcript abundance in only one zooid but not in any other zooid within a given species (Fig. \@ref(fig:unique-zooids), supplementary file 2).

For each species and zooid, we found GO term enrichment for biological processes consistent with the functional specialization of the zooid. In gastrozooids that are solely responsible for feeding and digestion, for example, we found these zooids to be enriched for genes involved in chitin, glutathione, and peptide catabolism, proteolysis, as well as metabolism of carbohydrates. Likewise for male gonodendra, we found GO term enrichment for biological processes such as sperm flagellum, mitotic cell cycle process, DNA replication. For female gonodendra, there was GO term enrichment for a number of biological processes including mitotic cell cycle process, DNA replication (in *Agalma elegans*), as well as a number of signalling pathways and developmental processes (in *Frillagalma vityazi*).

@@

In *Physalia physalis* `r dge_sig[[6]]$TenpalmatGasmat %>% nrow()` genes showed higher transcript abundance in the mature tentacular palpon, compared to `r dge_sig[[6]]$GasmatTenpalmat %>% nrow()` genes in the mature gastrozooid. A number of genes had higher transcript abundance in the mature tentacular palpon relative to all other tissues, of which, `r Physalia_Tenpalmat_unique %>% length()` genes were upregulated relative to other tissues that were not shared with the gastrozooid. In the gastrozooid `r Physalia_Gasmat_unique %>% length()` genes were upregulated relative to other tissues that were not shared with the tentacular palpon. A number of genes upregulated in the tentacular palpon are uncharacterized, however we identified 46 putative toxins in the tentacular palpon, including hemostasis interfering and platelet aggregation activating toxins, phospholipases, serineproteases, hydrolases, metalloendoproteases, calglandulin-like genes, and a neurotoxin. By contrast, in the gastrozooid, we found 59 upregulated putative toxins, including pore-forming Conoporin-Cn1-like and Tereporin-Ca1-like toxins, multiple neurotoxins, hydrolases, serine proteases, toxins likely involved in the promotion of blood coagulation and inhibition of platelet aggregation. Reflecting the role of the gastrozooid in digestion, we also find upregulation of digestive enzymes, including chymotrypsin-like genes.

Between the white mature gastrozooid and the yellow mature gastrozooid in *Bargmannia elongata*, few upregulated genes were identified (`r dge_sig[[2]]$GaswhimatGasyelmat %>% nrow()` genes were up in "white" mature gastrozooids relative to `r dge_sig[[2]]$GasyelmatGaswhimat %>% nrow()` genes in "yellow" gastrozooids). Among genes that were significantly upregulated in either "white" or "yellow" gastrozooids relative to all other tissues, `r Bargmannia_Gasyelmat_unique %>% length()` genes were unique to "yellow" gastrozooids and not found in "white" gastrozooids, and `r Bargmannia_Gaswhimat_unique %>% length()` genes were found in "white" gastrozooids and not found in "yellow" gastrozooids.

Finally, in *Agalma elegans* very few significantly upregulated genes were identified between the B palpon and gastric palpons  (`r dge_sig[[1]]$PalBmatPalmat %>% nrow()` had higher transcript abundance in B palpons and `r dge_sig[[1]]$PalmatPalBmat %>% nrow()` had higher transcript abundance in gastric palpons. All three genes have no significant blast hit and did not map to any gene trees). Genes were identified that are upregulated in B palpons relative to all other zooids (gastric palpons were excluded). Of this, `r Agalma_PalponB_unique %>% length()` genes were differentially expressed in the B palpons. Most of these genes overlapped with those differentially expressed in the gastric palpons relative to all other zooids (`r intersect(Agalma_PalponB_unique,Agalma_Palpon_unique) %>% length()` genes).

>>>>

readr::write_csv(edges_tidy_directional_export, "Supplementary_Data/Supplementaryfile9.csv", na = "NA")

@@

# Evolution of gene expression across species and specialized zooids in Siphonophora {-}
Catriona Munro^1,2^,  Stefan Siebert^3^, Mark Howison^4^, Felipe Zapata^5^, Casey W. Dunn^6^
\newline
\newline
^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI 02912, USA \newline
^2^ Current address: Collège de France, PSL Research University, CNRS, Inserm, Center for Interdisciplinary Research in Biology, 75005 Paris, France \newline
^3^ Department of Molecular and Cellular Biology, University of California, Davis, California, 95616, USA \newline
^4^ Research Improving People's Lives (RIPL), Providence, RI, USA \newline
^5^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA 90095, USA \newline
^6^ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT 06520, USA

@@

Siphonophores are complex colonial organisms, consisting of asexually-produced bodies (called zooids) that share a gastrovascular cavity but that are functionally specialized to specific tasks, including feeding, swimming, and reproducing. Though this extreme functional specialization has captivated biologists for generations, the genomic underpinnings of this phenomenon remain poorly understood. We use RNA-seq to investigate gene expression patterns in five zooids, and one specialized tissue (the pneumatophore, a gas filled float), across seven siphonophore species. Analyses of gene expression across genes and species present several challenges, which we address, in particular the identification of comparable homologous genes and appropriate normalization of expression across species.

@@

Within species, we identified hundreds of zooid-specific genes, as well as a number of putative transcription factors showing differential expression in particular zooids and developmental stages. To compare expression across species, we introduce a novel expression metric and new analytical approaches. We found that gene expression patterns tended to be largely consistent in zooids with the same function across species, but also found evidence of large lineage specific shifts in gene expression.

@@

Differential gene expression analyses enable identification of a large number of genes with zooid-specific expression, however comparisons across species are limited in scope and largely qualitative. Classical comparative analyses enable the identification of species/zooid-specific expression patterns within orthologs. Using branch filtering methods, we extended analyses to a larger number of genes, and identified instances of lineage-specific expression, as well as shared expression within zooids and across species.

## Background {-}

@@

In this study we use RNA-seq data to investigate the functional specialization and evolution of zooids across siphonophore species. To carry out these analyses, we also address three key challenges to study the evolution of gene expression in a comparative framework [@dunn2013phylogenetic]. First, we introduce a new metric to normalize gene expression which is valid for comparison across species. A common metric used to quantify gene expression is Transcripts Per Million (TPM) [@li2011rsem; @wagner2012measurement]. TPM is a relative measure of expression which depends on the number of genes present in a reference. Hence, TPM values are a valid metric when comparing libraries within a single species, but are not directly comparable across species when the references are incomplete, or genes have been gained and lost in the course of evolution. To address this issue, we introduce Transcripts Per Million 10K (TPM10K), a metric which normalizes TPM to account for different sequencing depths among species (see Methods for details).

Second, we further account for gene and species effects to compare gene expression across genes and species. Normalized read counts produced by RNA-seq experiments are proportional to gene expression, but they are also impacted by factors that can vary across genes and species such as gene sequence and length. Therefore, to estimate and compare expected gene counts it is necessary to model such factors with unknown species- and gene-specific counting-efficiency coefficients [see @dunn2013phylogenetic]. A direct comparison of expected gene counts among species, however, can be misleading if differences in counts are simply due to differences in counting efficiency and not due to differences in expression. To address this issue, we use ratios of expected counts to compare expression across genes and species. Using ratios of counts, we are able to eliminate species- and gene- specific counting efficiency within species, as the unknown counting efficiency factor is in both the numerator and the denominator, and is thus removed prior to comparisons among species [@dunn2013phylogenetic].

Third, we use a novel approach to investigate the evolution of gene expression on gene trees with complex evolutionary histories. Most evolutionary studies of gene expression focus exclusively on comparing expression values of strict orthologs (i.e., gene lineages related only by speciation events) which are shared across all species. We call this type of classical analysis Speciation Tree Filtering (STF) [e.g @Brawand:2011du; @cardoso2019gene; @Levin:2016bp; @yang2013organ]. This limits analyses to the subset of genes that have no evidence of duplication. This is an all or nothing approach, where gene trees or subtrees (clades) are retained if they have one sequence for each species, and discarded if there are multiple sequences in any of the species. Because this is usually such a small fraction of genes, traditional STF analyses discard a large fraction of the data, including many gene families that may be of interest to the investigator.

Because the history of genes is characterized by more complex scenarios involving gene duplication and speciation, we introduce a new method that takes advantage of this rich history to examine the evolution of gene expression across more genes and species. We call this method Speciation Branch Filtering (SBF) (Fig. \@ref(fig:methods-comp)), as we map expression data to gene phylogenies and identify equivalent branches in the species tree which are descended from speciation events in order to make valid comparisons across species. While STF is a gene tree filtering approach, where entire trees or subtrees are discarded due to even a single duplication event, SBF is a branch filtering approach. This means that SBF retains many informative branches from trees that would be removed by STF, and preserves many more evolutionary comparisons for analyses.

Briefly, the principles of SBF are as follows: first we identify speciation and duplication events in a given gene tree, specifically labelling nodes that correspond to speciation events in the species tree (Fig. \@ref(fig:methods-comp), step 1). Next, we map gene expression values to the tips of the gene tree (Fig. \@ref(fig:methods-comp), step 2), and use phylogenetic methods to reconstruct expression values at the internal nodes (Fig. \@ref(fig:methods-comp), step 3). Expression values are mapped and reconstructed for each zooid/tissue separately, with the assumption that the structure is homologous across species. Then, we calculate scaled expression values across branches, by subtracting the expression value at the child node from the parent node, and dividing by branch length (branch lengths are calibrated against branches in the species tree) (Fig. \@ref(fig:methods-comp), step 4). Finally, we identify branches in the gene trees that correspond to equivalent branches in the species tree (hereafter species-equivalent branches) (Fig. \@ref(fig:methods-comp), step 5). Species-equivalent branches are gene tree branches with parent and child nodes that are both speciation events and that correspond to branches in the species tree. Each branch in the species tree is given a unique identifier (i.e., a letter) and the species-equivalent branches in gene trees are given the same identifier (Fig. \@ref(fig:methods-comp), step 5). This method enables the selection of branches within gene trees that are equivalent to branches within the species tree, and are thus comparable with one another across all gene trees. Unlike the STF approach, this approach considers equivalent branches that are descended from speciation events, but that have more complex evolutionary histories. For example, due to deeper gene duplication events, gene trees often contain multiple branches that correspond to the same branch in the species tree (Fig. \@ref(fig:methods-comp), step 5). Our method allows us to consider all of these branches.

Using the methods described above, combined with classical differential gene expression (DGE) analysis, we conducted DGE analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. Second, we examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Third we compared gene expression across the seven siphonophore species, to identify putative clade/lineage or zooid specific expression patterns.

(ref:methods-comp) New phylogenetic approach to identify evolutionary changes in gene expression, called Speciation Branch Filtering (SBF). Step 1, we label each of the nodes in the species tree, and identify equivalent speciation nodes across every gene tree (an exemplar is shown here). Step 2, we map expression values (TPM10K) to the tips (expression values are mapped and reconstructed for each homologous zooid separately). Step 3, we reconstruct ancestral trait expression values at all internal nodes where expression data are available. Step 4, we calculate scaled change in gene expression (child node expression - parent node expression / branch length). Branch length is calibrated to the species tree branch lengths. Step 5, we identify branches in gene trees that correspond to equivalent branches in the species tree. There may be more than one branch in a gene tree that corresponds to the same branch in the species tree.

@@

# Bargmannia_GO_Gasmat<- Go_lookup("Bargmannia",Bargmannia_Gaswhimat) %>% add_column(Species="Bargmannia elongata") %>%  add_column(zooid="Gastrozooid mature")
# 
# Agalma_GO_Gasmat<- Go_lookup("Agalma",Agalma_gasmat) %>% add_column(Species="Agalma elegans") %>%  add_column(zooid="Gastrozooid mature")
# 
# Nanomia_GO_Gasmat<- Go_lookup("Nanomia",Nanomia_Gasmat)  %>% add_column(Species="Nanomia bijuga") %>%  add_column(zooid="Gastrozooid mature")
# 
# Frillagalma_GO_Gasmat<- Go_lookup("Frillagalma",Frillagalma_Gasmat) %>% add_column(Species="Frillagalma vityazi") %>%  add_column(zooid="Gastrozooid mature")
# 
# Physalia_GO_Gasmat<- Go_lookup("Physalia",Physalia_Gasmat) %>% add_column(Species="Physalia physalis") %>%  add_column(zooid="Gastrozooid mature")
# 
# Physalia_GO_Tenpalmat<- Go_lookup("Physalia",Physalia_Tenpalmat) %>% add_column(Species="Physalia physalis") %>%  add_column(zooid="Tentacular palpon mature")
# 
# Bargmannia_GO_Gasdev<- Go_lookup("Bargmannia",Bargmannia_Gasdev) %>% add_column(Species="Bargmannia elongata") %>%  add_column(zooid="Gastrozooid developing") 
# 
# Nanomia_GO_Gasdev<- Go_lookup("Nanomia",Nanomia_Gasdev) %>% add_column(Species="Nanomia bijuga") %>%  add_column(zooid="Gastrozooid developing") 
# 
# Frillagalma_GO_Gasdev<- Go_lookup("Frillagalma",Frillagalma_Gasdev) %>% add_column(Species="Frillagalma vityazi") %>%  add_column(zooid="Gastrozooid developing") 
# 
# Physalia_GO_Gasdev<- Go_lookup("Physalia",Physalia_Gasdev) %>% add_column(Species="Physalia physalis") %>%  add_column(zooid="Gastrozooid developing") 
# 
# Bargmannia_GO_Gonmal<- Go_lookup("Bargmannia",Bargmannia_Gonmal) %>% add_column(Species="Bargmannia elongata") %>%  add_column(zooid="Gonodendron male") 
# 
# Frillagalma_GO_Gonmal<- Go_lookup("Frillagalma",Frillagalma_Gonmal)  %>% add_column(Species="Frillagalma vityazi") %>%  add_column(zooid="Gonodendron male") 
# 
# Agalma_GO_Gonmal<- Go_lookup("Agalma",Agalma_Gonmal)  %>% add_column(Species="Agalma elegans") %>%  add_column(zooid="Gonodendron male") 
#  
# Frillagalma_GO_Gonfem<- Go_lookup("Frillagalma",Frillagalma_Gonfem) %>% add_column(Species="Frillagalma vityazi") %>%  add_column(zooid="Gonodendron female") 
# 
# Agalma_GO_Gonfem<- Go_lookup("Agalma",Agalma_Gonfem) %>% add_column(Species="Agalma elegans") %>%  add_column(zooid="Gonodendron female") 
# 
# Nanomia_GO_Gonfem<- Go_lookup("Nanomia",Nanomia_Gonfem) %>% add_column(Species="Nanomia bijuga") %>%  add_column(zooid="Gonodendron female") 
# 
# Nanomia_GO_Palmat<- Go_lookup("Nanomia",Nanomia_Palmat) %>% add_column(Species="Nanomia bijuga") %>%  add_column(zooid="Palpon mature") 
# 
# Frillagalma_GO_Palmat<- Go_lookup("Frillagalma",Frillagalma_Palmat) %>% add_column(Species="Frillagalma vityazi") %>%  add_column(zooid="Palpon mature") 
# 
# Agalma_GO_Palmat<- Go_lookup("Agalma", Agalma_Palpon) %>% add_column(Species="Agalma elegans") %>%  add_column(zooid="Palpon mature") 
# 
# Nanomia_GO_Necdev<- Go_lookup("Nanomia",Nanomia_Necdev) %>% add_column(Species="Nanomia bijuga") %>%  add_column(zooid="Nectophore developing") 
# 
# Frillagalma_GO_Necdev<- Go_lookup("Frillagalma",Frillagalma_Necdev) %>% add_column(Species="Frillagalma vityazi") %>%  add_column(zooid="Nectophore developing") 
# 
# Agalma_GO_Necdev<- Go_lookup("Agalma", Agalma_necdev) %>% add_column(Species="Agalma elegans") %>%  add_column(zooid="Nectophore developing") 
# 
# Bargmannia_GO_Necdev<-Go_lookup("Bargmannia", Bargmannia_necdev) %>% add_column(Species="Bargmannia elongata") %>%  add_column(zooid="Nectophore developing") 
#  
# Diphyes_GO_Gon<- Go_lookup("Diphyes",Diphyes_Gon) %>% add_column(Species="Diphyes dispar") %>%  add_column(zooid="Gonophore") 
# 
# All_GO<- bind_rows(Bargmannia_GO_Gasmat,Agalma_GO_Gasmat,Nanomia_GO_Gasmat,Frillagalma_GO_Gasmat,Physalia_GO_Gasmat,Physalia_GO_Tenpalmat,Bargmannia_GO_Gasdev,Nanomia_GO_Gasdev,Frillagalma_GO_Gasdev,Physalia_GO_Gasdev,Bargmannia_GO_Gonmal,Frillagalma_GO_Gonmal,Agalma_GO_Gonmal, Frillagalma_GO_Gonfem,Agalma_GO_Gonfem,Nanomia_GO_Gonfem,Nanomia_GO_Palmat,Frillagalma_GO_Palmat,Agalma_GO_Palmat,Nanomia_GO_Necdev,Frillagalma_GO_Necdev,Agalma_GO_Necdev,Bargmannia_GO_Necdev, Diphyes_GO_Gon)
# 
# All_GO <- All_GO %>% select("Species", "zooid","category","over_represented_pvalue","under_represented_pvalue","numDEInCat","numInCat","term","ontology","bh_adjust")
# 
# readr::write_csv(All_GO, "Supplementary_Data/Supplementaryfile3.csv", na = "NA")

@@

We sequenced mRNA from microdissected zooids from seven different siphonophore species (at least two replicate colonies) and mapped these short-read libraries to previously published transcriptomes [@munro2018improved]. We collected RNA-seq data from, where possible, 5 different zooids and one specialized tissue, the pneumatophore, as well as unique zooids specific to *Agalma elegans* (B palpons), *Physalia physalis* (tentacular palpon), and *Bargmannia elongata* (yellow and white gastrozooids), and where possible developing and mature zooids (Table \@ref(tab:summarize-libraries)). Due to species availability, we were not able to sample more than one replicate for some zooids - single replicates were excluded from downstream DGE analyses (see Fig. \@ref(fig:zooids-sampled)).

@@

The second component of variation we considered was biological variation among sampled colonies (Fig. \@ref(fig:pca-diphyes)--\@ref(fig:pca-physalia)). Specimens were collected in the wild at different depths and over different time periods, but despite these environmental factors there was remarkably little variation among sampled colonies. Some samples did show greater variation across replicates, such as a developing palpon replicate in *Nanomia bijuga* (Fig. \@ref(fig:pca-nanomia)), and a developing gastrozooid and male gonodendra in *Frillagalma vityazi* (Fig. \@ref(fig:pca-Frillagalma)).

The third component of variation we considered was among zooids/specialized tissues within species. This was the greatest component of variation (Fig. \@ref(fig:pca-diphyes)--\@ref(fig:pca-physalia)). We identified genes that significantly differentially expressed in particular zooids, and in the pneumatophore, for each of the sampled species, based on pairwise comparisons (supplementary file 1). In each case, significally differentially expressed refers to genes with higher transcript abundance relative to the other zooid. As it is possible for the same gene to be significantly differentially expressed in pairwise comparisons of several zooids , we also identified a subset of genes that are significantly differentially expressed in only one zooid but not in any other zooid within a given species (Fig. \@ref(fig:unique-zooids), supplementary file 2).

For each species and zooid, we found GO term enrichment for biological processes consistent with the functional specialization of the zooid (supplementary file 3). In gastrozooids that are solely responsible for feeding and digestion, for example, we found these zooids to be enriched for genes involved in chitin, glutathione, and peptide catabolism, proteolysis, as well as metabolism of carbohydrates. Likewise for male gonodendra, we found GO term enrichment for biological processes such as sperm flagellum, mitotic cell cycle process, DNA replication. For female gonodendra, there was GO term enrichment for a number of biological processes including mitotic cell cycle process, DNA replication (in *Agalma elegans*), as well as a number of signalling pathways and developmental processes (in *Frillagalma vityazi*).

@@

In *Physalia physalis* `r dge_sig[[6]]$TenpalmatGasmat %>% nrow()` genes showed significant differential expression (in all cases, higher transcript abundance relative to the other zooid) in the mature tentacular palpon, compared to `r dge_sig[[6]]$GasmatTenpalmat %>% nrow()` genes in the mature gastrozooid (supplementary file 4). A number of genes were significantly differentially expressed in the mature tentacular palpon relative to all other tissues, of which, `r Physalia_Tenpalmat_unique %>% length()` genes were significantly differentially expressed relative to other tissues that were not shared with the gastrozooid. In the gastrozooid `r Physalia_Gasmat_unique %>% length()` genes were significantly differentially expressed relative to other tissues that were not shared with the tentacular palpon. A number of genes significantly differentially expressed in the tentacular palpon are uncharacterized, however we identified 46 putative toxins in the tentacular palpon, including hemostasis interfering and platelet aggregation activating toxins, phospholipases, serineproteases, hydrolases, metalloendoproteases, calglandulin-like genes, and a neurotoxin (supplementary file 5). By contrast, in the gastrozooid, we found 59 significantly differentially expressed putative toxins  (supplementary file 6), including pore-forming Conoporin-Cn1-like and Tereporin-Ca1-like toxins, multiple neurotoxins, hydrolases, serine proteases, toxins likely involved in the promotion of blood coagulation and inhibition of platelet aggregation. Reflecting the role of the gastrozooid in digestion, we also find significant differential expression of digestive enzymes, including chymotrypsin-like genes.

Between the white mature gastrozooid and the yellow mature gastrozooid in *Bargmannia elongata*, few significantly differentially expressed genes were identified (`r dge_sig[[2]]$GaswhimatGasyelmat %>% nrow()` genes were up in "white" mature gastrozooids relative to `r dge_sig[[2]]$GasyelmatGaswhimat %>% nrow()` genes up in "yellow" gastrozooids) (supplementary file 4). Among genes that were significantly differentially expressed in either "white" or "yellow" gastrozooids relative to all other tissues, `r Bargmannia_Gasyelmat_unique %>% length()` genes were unique to "yellow" gastrozooids and not found in "white" gastrozooids, and `r Bargmannia_Gaswhimat_unique %>% length()` genes were found in "white" gastrozooids and not found in "yellow" gastrozooids.

Finally, in *Agalma elegans* very few significantly differentially expressed genes were identified between the B palpon and gastric palpons  (`r dge_sig[[1]]$PalBmatPalmat %>% nrow()` were significantly differentially expressed in B palpons and `r dge_sig[[1]]$PalmatPalBmat %>% nrow()` were significantly differentially expressed in gastric palpons (supplementary file 4). All three genes have no significant blast hit and did not map to any gene trees). Genes were identified that are significantly differentially expressed in B palpons relative to all other zooids (gastric palpons were excluded). Of this, `r Agalma_PalponB_unique %>% length()` genes were differentially expressed in the B palpons. Most of these genes overlapped with those differentially expressed in the gastric palpons relative to all other zooids (`r intersect(Agalma_PalponB_unique,Agalma_Palpon_unique) %>% length()` genes).

```{r uniqe-zooid-supp, verbose= FALSE, message = FALSE, echo=FALSE, comment=NA, warning=FALSE}

TentacularDE<-dge_sig[[6]]$TenpalmatGasmat %>% add_column(Species="Physalia physalis") %>% add_column(PairwiseComparison="Tentacular palpon/Gastrozooid")

GastroDE<- dge_sig[[6]]$GasmatTenpalmat %>% add_column(Species="Physalia physalis") %>% add_column(PairwiseComparison="Gastrozooid/Tentacular palpon")

GaswhimatDE <- dge_sig[[2]]$GaswhimatGasyelmat %>% add_column(Species="Bargmannia elongata") %>% add_column(PairwiseComparison="Gastrozooid white/Gastrozooid yellow")
